\pkg{mlr3spatiotempcv}: Spatiotemporal Resampling Methods for \pkg{mlr3}

### 2021-04-04 paper/paper.Rmd
first readable draft

---
documentclass: jss
author:
  - name: Patrick Schratz
    affiliation: 'Friedrich-Schiller-University Jena \AND'
    # affiliaton2: 'Ludwig-Maximilians-University Munich'
    # use this syntax to add text on several lines
    address: |
      | Faculty of Chemistry and Earth Sciences
      | Department of Geography
      | GIScience group
    email: \email{patrick.schratz@uni-jena.de}
  - name: Marc Becker
    affiliation: 'Ludwig-Maximilians-University Munich'
    address: |
      Department of Statistics
      Statistical Learning and Data Science group
    # To add another line, use \AND at the end of the previous one as above
  - name: Michel Lang
    affiliation: TU Dortmund University
    address: |
      | Faculty of Statistics
title:
  formatted: "\\pkg{mlr3spatiotempcv}: Spatiotemporal Resampling Methods for \\pkg{mlr3}"
  # If you use tex in the formatted title, also supply version without
  plain:     "mlr3spatiotempcv: Spatiotemporal Resampling Methods for mlr3"
  # For running headers, if needed
  # short:     "\\pkg{foo}: A Capitalized Title"
abstract: >
  The R package [mlr3spatiotempcv](https://mlr3spatiotempcv.mlr-org.com) is an extension package of the machine learning framework [mlr3](https://mlr3.mlr-org.com) [@mlr3].
  It adds support for spatiotemporal resampling methods and respective visualization methods.
  Spatiotemporal resampling methods are needed to account for the existence of different autocorrelation types (spatial autocorrelation (SAC), temporal autocorrelation (TAC) or spatio-temporal autocorrelation (STAC)) in datasets.
  These autocorrelation types cause biased performance estimates of varying magnitude in cross-validation when ignored.
  At the time of writing, various R packages implementing different spatiotemporal partitioning strategies exist: CAST [@cast], blockCV [@blockCV] or sperrorest [@sperrorest].
  There are some resampling methods used in scientific studies for which the source code is either not available or a different programming language has been used [@roberts2017; @pohjankukka2017]

@@

In the last years awareness for spatiotemporal autocorrelation in ML tasks has grown [@pohjankukka2017; @roberts2017].
Subsequently, scientists started writing code to account for the issue in their studies.
With R being one of the most used languages with respect to modeling/statistics and academia in general, R packages were developed aiming to simplify the data splitting with respect to methods able to account for spatial autocorrelation [@cast, @blockCV].
FIXME: adds refs
This is a typical process for the R community as writing R packages has been made relatively straightforward in recent years thanks to efforts from various people/organizations such as Bioconductor, CRAN, ropensci or the tidyverse [@cran, @ropensci, @tidyverse]
Also the effort of bundling code into a defined package structure makes it easier for others to re-use it, which is a good effort towards reproducible science.

The downside of this culture is that "island solutions" are being created, i.e. small packages which serve a particular (sometimes very niche) use case.
Usually the "Do one thing and do it well" UNIX philosophy is widely known and used in practice [@raymond2003art].
However, the issue with R packages is that not every package is necessarily easily compatiable with other ones due to differing input and returns objects, making it problematic to write a clean and tidy workflow.
Also many R packages are written by single persons which not necessarly have a complete overview over existing implementations.
These points may lead to the following possible issues in practice:

- Different syntax.
- Support for only some predictor types (e.g. numeric and factor features but no ordered factors).
- Different return values.
- Different namings for the same method.
- Unclear longterm maintenance.

These points add substantial overhead and/or reproducibility issues for anyone who wants to make use of the provided implementations within their studies.
\pkg{mlr3spatiotempcv} aims to provide a solution for some of the points outlined above by wrapping all spatiotemporal resampling methods available in \proglang{R} packages.
By embedding the package within the \pkg{mlr3} ecosystem and following its unified syntax and scalable approach, \pkg{mlr3spatiotempcv} aims to provide future users longterm support for spatiotemporal partitioning methods in R.

@@

# Visualization of Spatiotemporal Partitions

@@

Unless specified by the user, the coordinate reference system (CRS) defaults to EPSG code 4326 (WGS84).
This is because a lat/lon based CRS is better suited for plotting purposes than a Mercator (UTM) one.
Note that setting the correct CRS for the given data *during construction* is very important.
Even though EPSG 4326 is a good fallback and often used for visualization purposes, spatial offsets of up to multiple meters may occur if the wrong CRS was passed initially.

Note: The following example uses a built-in task via function `tsk()`.
In practice however, one needs to create a spatiotemporal task first via `TaskClassifST()`/`TaskRegrST()`..

The spatial grouping of the k-means based approach (`"spcv_coords"`) (Fig.\@ref(fig:vis-spcv)) contrasts visually very well to the NSpCV (random) partitioning (Fig.\@ref(fig:vis-nspcv)).

```{r vis-spcv, fig.cap="Visualization of spatial partitioning", fig.pos="t!", fig.height=6, fig.width=8, cache=TRUE}
library(mlr3)
library(mlr3spatiotempcv)

@@

resampling_sp <- rsmp("repeated_spcv_coords", folds = 4, repeats = 2)
resampling_sp$instantiate(task)

autoplot(resampling_sp, task, fold_id = c(1:2), crs = 4326, size = 0.8) *
  ggplot2::scale_y_continuous(breaks = seq(-3.97, -4, -0.01)) *
  ggplot2::scale_x_continuous(breaks = seq(-79.06, -79.08, -0.01))
```

```{r vis-nspcv, fig.cap="Visualization of random partitioning", fig.pos="t!", fig.height=6, fig.width=8, cache=T}
resampling_nsp <- rsmp("repeated_cv", folds = 4, repeats = 2)
resampling_nsp$instantiate(task)

autoplot(resampling_nsp, task, fold_id = c(1:2), crs = 4326, size = 0.8) *
  ggplot2::scale_y_continuous(breaks = seq(-3.97, -4, -0.01)) *
  ggplot2::scale_x_continuous(breaks = seq(-79.06, -79.08, -0.01))
```

# Spatiotemporal Partitioning Methods

At the time of writing, the following spatiotemporal resampling methods (Table \@ref(tab:sptcv-methods)) were available across different R packages.
This list might increase over time and it is recommended to consult the homepage of \pkg{mlr3spatiotempcv} for an up-to-date list.

A low-level introduction is given for every method included at this point of this article.
For a discussion with respect to usage see section FIXME.

## Spatial Buffering

@@

```{r buffer, fig.cap="Visualization of 'Spatial buffering' method.", out.width="50%", cache=T}
resampling_buffer <- rsmp("spcv_buffer", theRange = 1000)
resampling_buffer$instantiate(task)

autoplot(resampling_buffer,
  size = 0.8, task = task, fold_id = 1, crs = 4326) *
  ggplot2::scale_y_continuous(breaks = seq(-3.97, -4, -0.01)) *
  ggplot2::scale_x_continuous(breaks = seq(-79.06, -79.08, -0.01))
```

@@

## Spatial Blocking

@@

The way how the blocks are created can be manifold:

1. Supplying a numeric value in meters via argument `theRange` (named just `range` in {mlr3spatiotempcv}) will create a quadratic block pattern across the study area.
  The aggregation of zones can either be random (default) (`selection = "random"`), systematic (`selection = "systematic"`) or checkerboard structure (`selection = "checkerboard"`).

@@

```{r block-random, cache = TRUE, fig.cap="Spatial blocking with option `selection = 'random'`.", out.width="50%", fig.pos="h"}
resampling_block_random <- rsmp("spcv_block", range = 1000, folds = 5)

autoplot(resampling_block_random,
  size = 0.8, fold_id = 1, task = task, crs = 4326,
  show_blocks = TRUE, show_labels = TRUE) *
  ggplot2::scale_y_continuous(breaks = seq(-3.97, -4, -0.01)) *
  ggplot2::scale_x_continuous(breaks = seq(-79.06, -79.08, -0.01))
```

@@

```{r block-systematic, cache=TRUE, fig.cap="Spatial blocking with option `selection = 'systematic'`.", out.width="50%", fig.pos="h"}
resampling_block_systematic <- rsmp("spcv_block",
  range = 1000, folds = 5,
  selection = "systematic"
)

autoplot(resampling_block_systematic,
  size = 0.8, fold_id = 1, task = task, crs = 4326,
  show_blocks = TRUE, show_labels = TRUE) *
  ggplot2::scale_y_continuous(breaks = seq(-3.97, -4, -0.01)) *
  ggplot2::scale_x_continuous(breaks = seq(-79.06, -79.08, -0.01))
```

@@

```{r block-checkerboard, cache=TRUE, fig.cap="Spatial blocking with option `selection = 'checkerboard'`.", out.width="50%", fig.pos="h"}
resampling_block_checkerboard <- rsmp("spcv_block",
  range = 1000, folds = 5,
  selection = "checkerboard")

autoplot(resampling_block_checkerboard,
  size = 0.8, fold_id = 1, task = task, crs = 4326,
  show_blocks = TRUE, show_labels = TRUE) *
  ggplot2::scale_y_continuous(breaks = seq(-3.97, -4, -0.01)) *
  ggplot2::scale_x_continuous(breaks = seq(-79.06, -79.08, -0.01))
```

@@

```{r env-1, fig.cap="Environmental blocking using a single predictor.", cache=TRUE, out.width="50%", fig.pos="h"}
resampling_env <- rsmp("spcv_env",
  features = "distdeforest", folds = 5)

autoplot(resampling_env,
  size = 0.8, fold_id = 1, task = task, crs = 4326,
  show_blocks = TRUE, show_labels = TRUE) *
  ggplot2::scale_y_continuous(breaks = seq(-3.97, -4, -0.01)) *
  ggplot2::scale_x_continuous(breaks = seq(-79.06, -79.08, -0.01))
```

@@

```{r env-2, fig.cap="Environmental blocking using multiple predictors.", cache=TRUE, out.width="50%", fig.pos="h"}
resampling_env_multi <- rsmp("spcv_env",
  features = c("distdeforest", "slope"), folds = 5)

autoplot(resampling_env_multi,
  size = 0.8, fold_id = 1, task = task, crs = 4326,
  show_blocks = TRUE, show_labels = TRUE) *
  ggplot2::scale_y_continuous(breaks = seq(-3.97, -4, -0.01)) *
  ggplot2::scale_x_continuous(breaks = seq(-79.06, -79.08, -0.01))
```

@@

```{r coords, fig.cap="Method 'Spatial CV' showing a k-means clustering approach based on coordinates.", cache=TRUE, out.width="50%", fig.pos="h"}
resampling_coords <- rsmp("spcv_coords", folds = 5)$instantiate(task)

autoplot(resampling_coords,
  size = 0.8, fold_id = 1, task = task, crs = 4326) *
  ggplot2::scale_y_continuous(breaks = seq(-3.97, -4, -0.01)) *
  ggplot2::scale_x_continuous(breaks = seq(-79.06, -79.08, -0.01))
```

@@

autoplot(resampling_cstf_time,
  fold_id = 5, task = task_spt, crs = 4326, plot3D = TRUE
)
```

```{r lto, echo=FALSE, out.width="90%", fig.cap="Three-dimensional visualization of method `sptcv\\_cstf` and sub-method Leave-Time-Out (LTO). To support showcasing the temporal grouping, five folds and five temporal levels in variable `Date` were used in this example.", fig.pos="h"}
knitr::include_graphics("lto.png")
```

@@

autoplot(resampling_cstf_loc,
  fold_id = 5, task = task_spt, crs = 4326, plot3D = TRUE)
```

```{r llo, echo=FALSE, out.width="90%", fig.cap="Three-dimensional visualization of method 'sptcv\\_cstf' and sub-method 'Leave-Location-Out' (LLO). To support showcasing the spatial grouping, the viewing angle has been adjusted to a view from top. Points are stacked by temporal levels and only points from the same location are chosen to form a partition.", fig.pos="h"}
knitr::include_graphics("llo.png")
```

@@

autoplot(resampling_cstf_time_loc,
  fold_id = 4, task = task_spt, crs = 4326, plot3D = TRUE,
  show_omitted = TRUE)
```

```{r llto, echo=FALSE, out.width="90%", fig.cap="Three-dimensional visualization of method 'sptcv\\_cstf' and sub-method 'Leave-Location-And-Time-Out' (LLTO). The grey points show the points omitted by this method.", cache=TRUE, fig.pos="h"}
knitr::include_graphics("llto.png")
```

@@

# List of spatiotemporal resampling methods

Table \@ref(tab:sptcv-methods) lists all available spatiotemporal resampling methods available in \pkg{mlr3spatiotempcv}.
Column `Package` denotes the upstream package from which the method was incorporated.
The internal identifier for the respective method in \pkg{mlr3spatiotempcv} is shown in column `Notation`.

# Comparing Spatial and Non-Spatial Resampling

To show the effect of spatial autocorrelation in modeling, a spatial and non-spatial CV using \pkg{mlr3spatiotempcv} will be shown in the following.
For simplicity, one of the most used methods (spatial k-means clustering after @sperrorest) will be used.
Individual methods are discussed in FIXME.

The performance of a simple classification tree (`"classif.ranger"`) is evaluated on a random repeated cross-validation (`"repeated_cv"`) with four folds and two repetitions.
In reality, more fold (between five and ten) and a higher number of repetitions should be used.
The chosen evaluation measure is the "Area Under the ROC Curve" (`"classif.auc"`) which is a common evaluation measure for binary responses.
The CV will be executed twice with the only difference being the use of a spatial resampling object (initialized with `rsmp("repeated_spcv_coords"`) instead of a non-spatial one (`rsmp("repeated_cv"`).

## Data Preparation

The following code chunk first loads all necessary R packages and sets a low verbosity for this example.
A seed is set for deterministic execution.

Next the mlr3 task `"ecuador"`, which is included in \pkg{mlr3spatiotempcv}, is loaded via `tsk()`.
This function is a convenience wrapper around the underlying R6 construction method for a spatiotemporal classification task (`mlr3spatiotempcv::TaskClassifST$new()`).

@@

By setting the logging threshold, which can be controlled via the \pkg{lgr} package [@lgr], to `"warn"`, verbosity is reduced to keep the example tidy.

@@

task <- tsk("ecuador")

learner <- lrn("classif.ranger", predict_type = "prob")
```

### Non-Spatial Cross-Validation

Here, a four-fold, two times repeated non-spatial resampling structure is constructed and instantianeted via method `$instantiate()`.
Next, the created resampling object `resampling_nsp` is passed to function `resample()` together with the Task and Learner created earlier to execute the cross-validation.

Last, the results from the CV are aggregated using the measure `"classif.auc"`, which is the technical notation for the Area Under the Receivers Operating Curve (AUROC).

```{r rr-nspcv, cache=TRUE, cache.lazy = FALSE}
resampling_nsp <- rsmp("repeated_cv", folds = 4, repeats = 2)
resampling_nsp$instantiate(task)
rr_nsp <- resample(
  task = task, learner = learner,
  resampling = resampling_nsp
)

@@

## Spatial Cross-Validation

The same approach as for non-spatial resampling is used for spatial resampling, expect the method being `"repeated_spcv_coords"` instead of `"repeated_cv"`.

```{r rr-spcv, cache=TRUE, cache.lazy = FALSE}
resampling_sp <- rsmp("repeated_spcv_coords", folds = 4, repeats = 2)
resampling_sp$instantiate(task)
rr_sp <- resample(
  task = task, learner = learner,
  resampling = resampling_sp
)

@@

The resulting AUC using Non-Spatial Cross-Validation (NSpCV) is around 0.15 better compared to the Spatial Cross-Validation (SpCV).
The better performance score results from over-optimism in the estimation with NSpCV.
The actual predictive performance on new data is much more realistically reflected by the performance estimated via SpCV.
The magnitude of this difference is variable as it depends on the dataset, the magnitude of STAC and the learner itself.
Algorithms with a higher tendency of overfitting to the training set will have a large spread in such scenarios.

# Discussion

## Choosing a Resampling Method

@@

## A critical evaluation of all methods

Even though all methods have their pros and cons and no universal method exists, this chapter should provide a critial view on the limitations of each method.

<!-- spcv coords: mention generic name -->

<!-- sptcv LLTO: discards many observations -->
<!-- sptcv LLO: requires additional variable to specify groupings -->
<!-- sptcv LTO: ignores spatial autocor entirely -->

<!-- spcv buffer: discards observations, LOO very expensive -->

<!-- missing: non-LOO method with buffer support -->

<!-- env block: hard to decide which variable to use for clustering -->
<!-- env block: more than one variable may lead to unforeseeable results -->

>>>>

---
documentclass: jss
author:
  - name: Patrick Schratz
    affiliation: 'Friedrich-Schiller-University Jena \AND'
    # affiliaton2: 'Ludwig-Maximilians-University Munich'
    # use this syntax to add text on several lines
    address: |
      | Faculty of Chemistry and Earth Sciences
      | Department of Geography
      | GIScience group
    email: \email{patrick.schratz@uni-jena.de}
  - name: Marc Becker
    affiliation: 'Ludwig-Maximilians-University Munich'
    address: |
      Department of Statistics
      Statistical Learning and Data Science group
    # To add another line, use \AND at the end of the previous one as above
  - name: Michel Lang
    affiliation: TU Dortmund University
    address: |
      | Faculty of Statistics
title:
  formatted: "\\pkg{mlr3spatiotempcv}: Spatiotemporal resampling methods for \\pkg{mlr3}"
  # If you use tex in the formatted title, also supply version without
  plain:     "mlr3spatiotempcv: spatiotemporal resampling methods for mlr3"
  # For running headers, if needed
  # short:     "\\pkg{foo}: A Capitalized Title"
abstract: >
  The R package [mlr3spatiotempcv](https://mlr3spatiotempcv.mlr-org.com) is an extension package of the machine learning framework [mlr3](https://mlr3.mlr-org.com) [@mlr3].
  It adds support for spatiotemporal resampling methods and respective visualization methods.
  Spatiotemporal resampling methods are needed to account for the existence of different autocorrelation types (spatial autocorrelation (SAC), temporal autocorrelation (TAC) or spatio-temporal autocorrelation (STAC)) in datasets.
  These autocorrelation types cause biased performance estimates of varying magnitude in cross-validation when ignored.
  At the time of writing, various R packages implementing different spatiotemporal partitioning strategies exist: CAST [@cast], blockCV [@blockCV] or sperrorest [@sperrorest].
  There are some resampling methods used in scientific studies for which the source code is either not available or a different programming language has been used [@roberts2017; @pohjankukka2017]

@@

In the last years awareness for spatiotemporal autocorrelation in ML tasks has grown [@pohjankukka2017; @roberts2017].
Subsequently, scientists started writing code to account for the issue in their studies.
With R being one of the most used languages with respect to modeling/statistics and academia in general, R packages were developed aiming to simplify the data splitting with respect to methods able to account for spatial autocorrelation [@cast, @blockCV].

This is a typical process for the R community as writing R packages has been made relatively straightforward in recent years thanks to efforts from various people/organizations such as Bioconductor, CRAN, ropensci or the tidyverse [@cran, @ropensci, @tidyverse]
Also the effort of bundling code into a defined package structure makes it easier for others to re-use it, which is a good effort towards reproducible science.

The downside which comes with this is that "island solutions" are being created, i.e. small packages which serve a particular (sometimes very niche) use case.
While the UNIX philosophy "Do one thing and do it well" [@raymond2003art] is widely known and used in the R community, R packages from different developers do often not play well with each other.
This is due to differing input and returns objects, requiring the need to pre-and postprocess objects around such R packages which provide tools for specific modeling tasks.
Also often R packages implement only an opinionated subset of methods which require users to combine multiple packages for different steps.

These points may lead to the following possible issues in practice:

- Differing syntax.
- Support for only some predictor types (e.g. numeric and factor features but no ordered factors).
- Differing return values.
- Differing namings for the same method.
- Unclear longterm maintenance.

These points add substantial overhead and/or reproducibility issues for anyone who wants to make use of the provided implementations within their studies.
\pkg{mlr3spatiotempcv} aims to provide a solution for some of the points outlined above by wrapping all spatiotemporal resampling methods available in \proglang{R} packages.
By embedding the package within the \pkg{mlr3} ecosystem, \pkg{mlr3spatiotempcv} aims to provide longterm support for spatiotemporal partitioning methods in R while making all additional benefits of the machine learning framework available to users.

@@

# Visualization of spatiotemporal partitions

@@

Visualization of partitioning methods is a vital part of \pkg{mlr3spatiotempcv} as most upstream packages lack support for such.
The semantics are presented first because plotting will be used extensively when presenting the integrated methods.

Unless specified by the user, the coordinate reference system (CRS) defaults to EPSG code 32717 (WGS84).
This is because a lat/lon based CRS is better suited for plotting purposes than a Mercator (UTM) one due tho a more descriptive axes labeling.
Setting an appropriate CRS for the given data *during task construction* is very important.
Even though EPSG 32717 is a good fallback and often used for visualization purposes, spatial offsets of up to multiple meters may occur if the wrong CRS was passed initially.

The following example creates a spatiotemporal task from scratch.
The data used here is also available as a built-in task via `tsk("ecuador")`.

Because the spatial grouping of the k-means based approach (`"spcv_coords"`) (Fig. \@ref(fig:vis-spcv)) contrasts visually very well to the NSpCV (random) partitioning (Fig. \@ref(fig:vis-nspcv)), this method was chosen to showcase the 2D plotting capabilities of \pkg{mlr3spatiotempcv}.

The partitions shown in the following are also used in the case study in section \@ref(sec:case-study).

Besides the required arguments of the \pkg{mlr3} superclass `Task`, `TaskClassifST` / `TaskRegrST` classes also require the following arguments passed via a named list `extra_args`:

- `coordinate_names`: Names of the features which represent the spatial coordinates.
  This is automatically inferred when a `sf` object is passed.
- `coords_as_features`: Whether the coordinates should be used as features during modeling.
  By default coordinates are not used as predictors.
- `crs`: The coordinate reference system of the data. Valid inputs are PROJ4 strings or EPSG codes in the format `ESPG: <code>`.

```{r create task, cache=TRUE}
library(mlr3)
library(mlr3spatiotempcv)

@@

# equal to `tsk("ecuador")`
b = mlr3::as_data_backend(ecuador)
b$hash = "_mlr3_tasks_ecuador_" # optional
task = TaskClassifST$new(
  id = "ecuador", b, target = "slides", positive = "TRUE",
  extra_args = list(
    coordinate_names = c("x", "y"), coords_as_features = FALSE,
    crs = "+proj=utm +zone=17 +south +datum=WGS84 +units=m +no_defs") # EPSG:32717
)
```

Due to space limitations only the first two folds are plotted.

```{r vis-spcv, fig.cap="Visualization of spatial partitioning", fig.pos="ht", fig.height=6, fig.width=8, cache=TRUE}
resampling_sp = rsmp("repeated_spcv_coords", folds = 4, repeats = 2)
resampling_sp$instantiate(task)

autoplot(resampling_sp, task, fold_id = c(1:2), crs = 32717, size = 0.8) *
  ggplot2::scale_y_continuous(breaks = seq(-3.97, -4, -0.01)) *
  ggplot2::scale_x_continuous(breaks = seq(-79.06, -79.08, -0.01))
```

```{r vis-nspcv, fig.cap="Visualization of random partitioning", fig.pos="ht", fig.height=6, fig.width=8, cache=T}
resampling_nsp = rsmp("repeated_cv", folds = 4, repeats = 2)
resampling_nsp$instantiate(task)

autoplot(resampling_nsp, task, fold_id = c(1:2), crs = 32717, size = 0.8) *
  ggplot2::scale_y_continuous(breaks = seq(-3.97, -4, -0.01)) *
  ggplot2::scale_x_continuous(breaks = seq(-79.06, -79.08, -0.01))
```

# Spatiotemporal partitioning methods

At the time of writing, the following spatiotemporal resampling methods (Tab. \@ref(tab:sptcv-methods)) were available across different R packages.
This list might increase over time and it is recommended to consult the homepage of \pkg{mlr3spatiotempcv} for an up-to-date list.
The internal identifier for the respective method in \pkg{mlr3spatiotempcv} is shown in column `Notation`.

A low-level introduction including visualization is given for every method in the following.
See section \@ref(sec:disc) for a discussion of the methods.

## Spatial buffering

@@

```{r buffer, fig.cap="Visualization of 'Spatial buffering' method.", out.width="50%", cache=T}
resampling_buffer = rsmp("spcv_buffer", theRange = 1000)
resampling_buffer$instantiate(task)

autoplot(resampling_buffer,
  size = 0.8, task = task, fold_id = 1, crs = 32717) *
  ggplot2::scale_y_continuous(breaks = seq(-3.97, -4, -0.01)) *
  ggplot2::scale_x_continuous(breaks = seq(-79.06, -79.08, -0.01))
```

@@

## Spatial blocking

@@

The way how the blocks are created can be manifold: Supplying a numeric value in meters via argument `theRange` (named just `range` in {mlr3spatiotempcv}) will create a quadratic block pattern across the study area.
The aggregation of zones can either be random (default) (`selection = "random"`), systematic (`selection = "systematic"`) or checkerboard structure (`selection = "checkerboard"`).

@@

```{r block-random, cache = TRUE, fig.cap="Spatial blocking with option `selection = 'random'`.", out.width="50%", fig.pos="ht"}
resampling_block_random = rsmp("spcv_block", range = 1000, folds = 5)

autoplot(resampling_block_random,
  size = 0.8, fold_id = 1, task = task, crs = 32717,
  show_blocks = TRUE, show_labels = TRUE) *
  ggplot2::scale_y_continuous(breaks = seq(-3.97, -4, -0.01)) *
  ggplot2::scale_x_continuous(breaks = seq(-79.06, -79.08, -0.01))
```

@@

```{r block-systematic, cache=TRUE, fig.cap="Spatial blocking with option `selection = 'systematic'`.", out.width="50%", fig.pos="ht"}
resampling_block_systematic = rsmp("spcv_block",
  range = 1000, folds = 5,
  selection = "systematic"
)

autoplot(resampling_block_systematic,
  size = 0.8, fold_id = 1, task = task, crs = 32717,
  show_blocks = TRUE, show_labels = TRUE) *
  ggplot2::scale_y_continuous(breaks = seq(-3.97, -4, -0.01)) *
  ggplot2::scale_x_continuous(breaks = seq(-79.06, -79.08, -0.01))
```

@@

```{r block-checkerboard, cache=TRUE, fig.cap="Spatial blocking with option `selection = 'checkerboard'`.", out.width="40%", fig.pos="h!"}
resampling_block_checkerboard = rsmp("spcv_block",
  range = 1000, folds = 5,
  selection = "checkerboard")

autoplot(resampling_block_checkerboard,
  size = 0.8, fold_id = 1, task = task, crs = 32717,
  show_blocks = TRUE, show_labels = TRUE) *
  ggplot2::scale_y_continuous(breaks = seq(-3.97, -4, -0.01)) *
  ggplot2::scale_x_continuous(breaks = seq(-79.06, -79.08, -0.01))
```

\newpage

### Rectangular blocks

Alternatively, arguments `rows` and `cols` can be specified instead of `theRange`.
This allows to create rectangular blocks which honor the supplied `rows` and `cols` settings.

```{r block-rows-and-cols, cache=TRUE, fig.cap="Spatial blocking with arguments `rows` and `cols`.", out.width="40%", fig.pos="ht"}
resampling_block_rows_cols = rsmp("spcv_block",
  rows = 6, cols = 3, folds = 5
)

autoplot(resampling_block_rows_cols,
  size = 0.8, fold_id = 1, task = task, crs = 32717,
  show_blocks = TRUE, show_labels = TRUE) *
  ggplot2::scale_y_continuous(breaks = seq(-3.97, -4, -0.01)) *
  ggplot2::scale_x_continuous(breaks = seq(-79.06, -79.08, -0.01))
```

\newpage

@@

```{r env-1, fig.cap="Environmental blocking using a single predictor.", cache=TRUE, out.width="50%", fig.pos="ht"}
resampling_env = rsmp("spcv_env",
  features = "distdeforest", folds = 5)

autoplot(resampling_env,
  size = 0.8, fold_id = 1, task = task, crs = 32717,
  show_blocks = TRUE, show_labels = TRUE) *
  ggplot2::scale_y_continuous(breaks = seq(-3.97, -4, -0.01)) *
  ggplot2::scale_x_continuous(breaks = seq(-79.06, -79.08, -0.01))
```

@@

```{r env-2, fig.cap="Environmental blocking using multiple predictors.", cache=TRUE, out.width="50%", fig.pos="ht"}
resampling_env_multi = rsmp("spcv_env",
  features = c("distdeforest", "slope"), folds = 5)

autoplot(resampling_env_multi,
  size = 0.8, fold_id = 1, task = task, crs = 32717,
  show_blocks = TRUE, show_labels = TRUE) *
  ggplot2::scale_y_continuous(breaks = seq(-3.97, -4, -0.01)) *
  ggplot2::scale_x_continuous(breaks = seq(-79.06, -79.08, -0.01))
```

@@

```{r coords, fig.cap="Method 'Spatial CV' showing a k-means clustering approach based on coordinates.", cache=TRUE, out.width="50%", fig.pos="ht"}
resampling_coords = rsmp("spcv_coords", folds = 5)$instantiate(task)

autoplot(resampling_coords,
  size = 0.8, fold_id = 1, task = task, crs = 32717) *
  ggplot2::scale_y_continuous(breaks = seq(-3.97, -4, -0.01)) *
  ggplot2::scale_x_continuous(breaks = seq(-79.06, -79.08, -0.01))
```

@@

autoplot(resampling_cstf_time,
  fold_id = 5, task = task_spt, crs = 32717, plot3D = TRUE
)
```

```{r lto, echo=FALSE, out.width="90%", fig.cap="Three-dimensional visualization of method `sptcv\\_cstf` and sub-method Leave-Time-Out (LTO). To support showcasing the temporal grouping, five folds and five temporal levels in variable `Date` were used in this example.", fig.pos="ht"}
knitr::include_graphics("lto.png")
```

@@

autoplot(resampling_cstf_loc,
  fold_id = 5, task = task_spt, crs = 32717, plot3D = TRUE)
```

```{r llo, echo=FALSE, out.width="90%", fig.cap="Three-dimensional visualization of method 'sptcv\\_cstf' and sub-method 'Leave-Location-Out' (LLO). To support showcasing the spatial grouping, the viewing angle has been adjusted to a view from top. Points are stacked by temporal levels and only points from the same location are chosen to form a partition.", fig.pos="ht"}
knitr::include_graphics("llo.png")
```

@@

autoplot(resampling_cstf_time_loc,
  fold_id = 4, task = task_spt, crs = 32717, plot3D = TRUE,
  show_omitted = TRUE)
```

```{r llto, echo=FALSE, out.width="90%", fig.cap="Three-dimensional visualization of method 'sptcv\\_cstf' and sub-method 'Leave-Location-And-Time-Out' (LLTO). The grey points show the points omitted by this method.", cache=TRUE, fig.pos="ht"}
knitr::include_graphics("llto.png")
```

@@

# Case study: Comparing spatial and non-spatial resampling {#sec:case-study}

To show the effect of spatial autocorrelation on performance estimates in cross-validation, a spatial and non-spatial CV using method `"spatial_coords"` will be shown in the following.

The performance of a simple classification tree (`"classif.ranger"`) is evaluated on a repeated cross-validation (`"repeated_cv"`) with four folds and two repetitions.
In reality, more fold (between five and ten) and a higher number of repetitions should be used.
The chosen evaluation measure is the "Area Under the ROC Curve" (`"classif.auc"`) which is a common evaluation measure for binary responses.
The CV will be executed twice with the only difference being the use of a spatial resampling object (initialized with `rsmp("repeated_spcv_coords"`) instead of a non-spatial one (`rsmp("repeated_cv"`).

## Data preparation

At first all necessary R packages are loaded and a lower verbosity is set to keep the output tidy.
Also a seed is set for deterministic execution.

Next the \pkg{mlr3spatiotempcv} built-in task `"ecuador"` is loaded via `tsk()`.

@@

@@

learner = lrn("classif.ranger", predict_type = "prob")
```

## Non-spatial cross-validation

A four-fold, two times repeated non-spatial resampling object is constructed and instantiated via method `$instantiate()`.
Next, the created resampling object `resampling_nsp` is passed to function `resample()` together with the Task and Learner created earlier to execute the cross-validation.

The results from the CV are then aggregated using the measure `"classif.auc"`, which is the package notation for the Area Under the Receivers Operating Curve (AUROC) measure.

```{r rr-nspcv, cache=TRUE, cache.lazy = FALSE}
resampling_nsp = rsmp("repeated_cv", folds = 4, repeats = 2)
resampling_nsp$instantiate(task)
rr_nsp = resample(
  task = task, learner = learner,
  resampling = resampling_nsp
)

@@

## Spatial cross-validation

The same approach as for non-spatial resampling is used for spatial resampling, expect replacing method `"repeated_cv"` by `"repeated_spcv_coords"`.

```{r rr-spcv, cache=TRUE, cache.lazy = FALSE}
resampling_sp = rsmp("repeated_spcv_coords", folds = 4, repeats = 2)
resampling_sp$instantiate(task)
rr_sp = resample(
  task = task, learner = learner,
  resampling = resampling_sp
)

@@

## Summary

The resulting AUC valiue of 0.77 using Non-Spatial Cross-Validation (NSpCV) is around 0.15 better compared to the Spatial Cross-Validation (SpCV) value of 0.61.
The better performance score results from over-optimism in the estimation with NSpCV.
The actual predictive performance on new data is much more realistically reflected by the performance estimated via SpCV.
The magnitude of this difference is variable as it depends on the dataset, the magnitude of STAC and the learner itself.
Algorithms with a higher tendency of overfitting to the training set will have a large spread in such scenarios.

The respective folds of the resamplings used in this example are shown in Fig. \@ref(fig:vis-spcv) and Fig. \@ref(fig:vis-nspcv).

# Discussion {#sec:disc}

## Choosing a resampling method

@@

## Comparing spatio-temporal resampling methods

<!-- high autocor on the border of test sets -->
<!-- sptcv LLTO: discards many observations -->
All methods which make use of all variables in the dataset suffer from the problem that on the border between train and test sets, both in the spatial and temporal domain, autocorrelation is quite high.
Only methods that omit some observations (`"spcv_buffer"`, `"sptcv_cstf"` with both `space_var` and `time_var` set) are able to reduce this problem.
However, these then suffer from the fact of potentially not being able to properly represent the dataset because only a subset of observations is used.

<!-- sptcv LLO & LTO: requires additional variable to specify groupings -->
Methods `"spcv_cstf"` (with only one of `space_var` or `time_var` set) require a variable in the dataset which should be used for grouping.
The specification of such is not always easy because the selection of such a variable also implies that the selected groups inherit substantial autorcorrelation and differ substantially from other groups.
Also, if the variable contains too many groups, the difference within train/test splits may become undesirably high and tend towards a LOO CV.

<!-- only some methods are suited for repeated execution -->
<!-- spcv_coords: limited variance due to limited k-means possibilties -->
For repeated executions of method `spcv_coords`, the variance between repetitions could be quite small due to the limited possibilities of different `k-means` clusters on the same dataset.
This applies even more to `"spcv_block"` with options `selection = "systematic"` and `selection = "checkerboard"` as the fold assignments are the same in each repetition.
Only methods with a random component (e.g. `"spcv_coords"` or `"spcv_block"` with option `selction = "random"`) may be suited for using multiple repetitions.

<!-- spcv buffer: discards observations, LOO very expensive -->
Method `"spcv_buffer"` makes use of the removal of observations to reduce the influence of spatial autocorrelation between train and test set.
This might be pretty effective in reducing autocorrelation but also requires to discard many variables and comes with the downsides of LOO-CV: many model fits which can be computationally very demanding and a high variance due to the high similarity of training datasets [@blockCV].

<!-- spcv coords: mention generic name -->
In @sperrorest, which is commonly used as the main reference for this k-means based spatial clustering approach, the method was not assigned a unique name.
This brings potential confusion for users and the community as the term "spatial CV" is used a generic one for the idea of accounting for spatial autocorrelation for spatial datasets.
Hence, in \pkg{mlr3spatiotempcv} this method was named `"spcv_coords"`, making use of the fact that the spatial coordinates are used for a cluster-based partition creation.
